---
title: "Advanced: Multi-trait, Enrichment, and Performance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced: Multi-trait, Enrichment, and Performance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dpi = 96
)
```

```{r setup}
library(gwasplot)
library(ggplot2)
data(example_gwas)
```

## Gene labels on peaks

By default, Manhattan plots label SNPs by their rs ID. In manuscripts you
typically want gene names instead. `manhattan_genes()` maps lead SNPs to
the nearest gene from a user-supplied table:

```{r gene-labels}
genes <- data.frame(
  chr = c(1, 2, 5, 7, 11, 15, 19, 22),
  start = c(2e6, 50e6, 80e6, 30e6, 60e6, 40e6, 10e6, 20e6),
  end = c(3e6, 52e6, 85e6, 35e6, 65e6, 45e6, 15e6, 25e6),
  gene = c("MTOR", "NRXN1", "TCF7L2", "CDKAL1", "KCNQ1", "FTO", "LDLR", "COMT")
)

manhattan_genes(example_gwas, genes = genes, gene_p_threshold = 0.001)
```

Use `arrow = TRUE` for a cleaner look when peaks are close together:

```{r gene-arrows}
manhattan_genes(example_gwas, genes = genes, gene_p_threshold = 0.001,
                arrow = TRUE, label_face = "italic")
```

## Top hits table

`top_hits()` extracts independent lead SNPs (window-based clumping),
adds a cytoband estimate, and optionally maps each to the nearest gene:

```{r top-hits}
top_hits(example_gwas, p_threshold = 0.001, genes = genes, n = 10)
```

The result is a plain data.frame — easy to export or paste into a manuscript.

## Highlighting genomic regions

Mark known regions (MHC, previously reported loci, CNVs) with a colored
band behind the Manhattan plot:

```{r regions}
plt <- manhattan_plot(example_gwas)

regions <- data.frame(
  chr = c(6, 1),
  start = c(25e6, 5e6),
  end = c(35e6, 30e6),
  label = c("MHC", "1p36")
)

highlight_regions(plt, regions, color = "#FFD700", alpha = 0.2)
```

## Multi-trait Manhattan

Compare multiple GWAS results on a single plot to identify shared loci
(pleiotropy):

```{r multitrait}
# Simulate a second trait
set.seed(123)
trait2 <- example_gwas
trait2$P <- runif(nrow(trait2))^3
trait2$P[sample(nrow(trait2), 10)] <- 10^(-runif(10, 5, 10))

multitrait_manhattan(
  BMI = example_gwas,
  Height = trait2,
  colors = "nature",
  highlight_shared = TRUE
)
```

### Customizing multi-trait plots

```{r multitrait-custom}
multitrait_manhattan(
  Trait1 = example_gwas,
  Trait2 = trait2,
  colors = c(Trait1 = "#E64B35", Trait2 = "#4DBBD5"),
  shapes = c(Trait1 = 16L, Trait2 = 17L),
  point_size = 1.0,
  alpha = 0.6,
  legend_position = "top"
)
```

## Enrichment Manhattan

Overlay functional genomic annotations to see if significant variants fall
in biologically relevant regions:

```{r enrichment}
# Define annotation regions
annotations <- data.frame(
  chr = c(1, 3, 5, 7, 11, 15, 19),
  start = c(5e6, 20e6, 50e6, 30e6, 60e6, 40e6, 10e6),
  end = c(30e6, 50e6, 120e6, 80e6, 100e6, 80e6, 40e6),
  category = c("Enhancer", "Gene", "eQTL", "Gene",
               "Enhancer", "eQTL", "Gene")
)

enrichment_manhattan(
  example_gwas,
  annotations = annotations,
  palette = "nature",
  label_top_n = 3
)
```

### Using a named list of annotations

```{r enrichment-list}
annot_list <- list(
  Promoters = data.frame(
    chr = c(1, 5, 10),
    start = c(2e6, 80e6, 50e6),
    end = c(10e6, 100e6, 70e6)
  ),
  Enhancers = data.frame(
    chr = c(3, 7, 15),
    start = c(30e6, 40e6, 20e6),
    end = c(60e6, 70e6, 50e6)
  )
)

enrichment_manhattan(
  example_gwas,
  annotations = annot_list,
  annotation_colors = c(Promoters = "#E64B35", Enhancers = "#4DBBD5")
)
```

## Circular Manhattan (multi-ring)

Compare multiple traits in a circos-style layout:

```{r circular-multi, fig.height=7}
circular_manhattan(
  list(BMI = example_gwas, Height = trait2),
  colors = "nature",
  point_size = 0.6
)
```

## Stratified QQ plot

Check for inflation in different subsets of your data:

```{r qq-stratified}
# Add MAF bins
example_gwas$MAF_bin <- cut(
  example_gwas$AF,
  breaks = c(0, 0.05, 0.2, 0.5),
  labels = c("Rare (<5%)", "Low (5-20%)", "Common (>20%)")
)

qq_plot(example_gwas, group = "MAF_bin")
```

## Performance: handling large datasets

gwasplot uses smart downsampling by default for datasets > 200k variants.
The algorithm:

1. **Keeps all significant SNPs** (p < 0.001)
2. **Keeps highlighted/labeled SNPs**
3. **Bins non-significant SNPs** by position and -log10(p)
4. **Deduplicates** within bins

This produces visually identical plots 10-100x faster:

```{r eval=FALSE}
# 10M SNPs — renders in seconds, not minutes
gwas_large <- read_gwas_table("large_gwas.txt")
manhattan_plot(gwas_large)  # automatic downsampling

# Disable if you need all points (e.g., for density analysis)
manhattan_plot(gwas_large, downsample = FALSE)

# Adjust target number of points
manhattan_plot(gwas_large, downsample_n = 500000)
```

## Combining plots for publication

Use `patchwork` to create complex multi-panel figures:

```{r patchwork, fig.height=8}
library(patchwork)

p1 <- manhattan_plot(example_gwas, label_top_n = 3) +
  labs(title = "A") + theme_nature()

p2 <- qq_plot(example_gwas) +
  labs(title = "B") + theme_nature()

p3 <- volcano_plot(example_gwas, label_top_n = 5) +
  labs(title = "C") + theme_nature()

p4 <- pvalue_heatmap(example_gwas, bin_size = 10e6) +
  labs(title = "D")

(p1 / (p2 | p3 | p4)) + plot_layout(heights = c(2, 1))
```

## Exporting for publication

```{r eval=FALSE}
# Nature single-column figure
fig <- manhattan_plot(example_gwas, label_top_n = 5) + theme_nature()
ggsave("figure1.pdf", fig, width = 89, height = 55, units = "mm")

# Multi-panel for supplementary
supp <- gwas_summary(example_gwas)
ggsave("supplementary_fig1.pdf", supp, width = 183, height = 200, units = "mm")
```
