---
title: "Customization: Themes, Palettes, and Presets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Customization: Themes, Palettes, and Presets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 96
)
```

```{r setup}
library(gwasplot)
library(ggplot2)
data(example_gwas)
```

## Color palettes

gwasplot includes 14 built-in palettes optimized for GWAS visualizations:

```{r palettes-list}
gwas_palettes()
```

### Using palettes

```{r palette-demo}
# Colorblind-safe (default for publication)
manhattan_plot(example_gwas, colors = gwas_palette("colorblind"))
```

```{r palette-nature}
# Nature-inspired
manhattan_plot(example_gwas, colors = gwas_palette("nature"))
```

```{r palette-brewer}
# RColorBrewer palettes
manhattan_plot(example_gwas, colors = gwas_palette("brewer_dark2"))
```

### Generating custom palettes

Use `gwas_palette()` with the `n` parameter to interpolate any number of colors:

```{r palette-custom}
# 5 colors from the nature palette
gwas_palette("nature", n = 5)

# Use with multi-trait plots
gwas_palette("science", n = 3)
```

## Journal themes

Pre-built themes matching journal style guides:

### Nature

```{r theme-nature}
manhattan_plot(example_gwas, label_top_n = 3) + theme_nature()
```

### Science

```{r theme-science}
manhattan_plot(example_gwas, label_top_n = 3) + theme_science()
```

### Cell

```{r theme-cell}
manhattan_plot(example_gwas, label_top_n = 3) + theme_cell()
```

### PLOS

```{r theme-plos}
manhattan_plot(example_gwas, label_top_n = 3) + theme_plos()
```

### Presentations and posters

```{r theme-presentation}
manhattan_plot(example_gwas) + theme_presentation()
```

```{r theme-poster}
manhattan_plot(example_gwas) + theme_poster()
```

## Publication presets

Presets combine a theme, color palette, and point sizes into one object:

```{r presets}
p <- gwas_preset("publication")
names(p)
```

Apply a preset:

```{r preset-apply}
manhattan_plot(
  example_gwas,
  colors = p$colors,
  point_size = p$point_size,
  label_top_n = 3
) + p$theme
```

Available presets:

| Preset | Use case | Font size | Point size |
|--------|----------|-----------|------------|
| `"publication"` | Journal figures | 7pt | 0.6 |
| `"presentation"` | Slides/talks | 16pt | 1.5 |
| `"poster"` | Conference posters | 20pt | 2.0 |
| `"exploratory"` | Data exploration | 11pt | 0.8 |

## Custom ggplot2 layers

Every gwasplot function returns a ggplot object. Add any ggplot2 layer:

```{r custom-layers}
manhattan_plot(example_gwas) +
  # Custom annotations
  annotate("rect", xmin = 1e8, xmax = 2e8, ymin = 0, ymax = Inf,
           alpha = 0.1, fill = "red") +
  # Custom labels
  labs(
    title = "Genome-wide Association Study",
    subtitle = "Heat stress tolerance in cattle",
    caption = "Data: Czech et al. (2026)"
  ) +
  theme_nature()
```

## Saving plots

Standard ggplot2 saving workflow:

```{r eval=FALSE}
p <- manhattan_plot(example_gwas) + theme_nature()

# For Nature (single column: 89mm)
ggsave("manhattan.pdf", p, width = 89, height = 60, units = "mm")

# For Science (two-column: 183mm)
ggsave("manhattan.pdf", p, width = 183, height = 100, units = "mm")

# High-res PNG for presentations
ggsave("manhattan.png", p, width = 10, height = 6, dpi = 300)
```
