---
title: "Getting Started with gwasplot"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with gwasplot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dpi = 96
)
```

## Installation

```{r eval=FALSE}
# Install from GitHub
pak::pak("bczech/gwasplot")
```

## Loading data

gwasplot can read GWAS results from multiple formats. The easiest way is
`read_gwas_table()` which auto-detects column names:

```{r}
library(gwasplot)

# Use bundled example data
data(example_gwas)
example_gwas
```

For real data, just point to your file:

```{r eval=FALSE}
# Auto-detect columns
gwas <- read_gwas_table("my_results.txt")

# Or use format-specific readers
gwas <- read_plink_assoc("results.assoc")
gwas <- read_regenie("results.regenie")
gwas <- read_gcta_mlma("results.mlma")
gwas <- read_gemma("results.assoc.txt")
```

You can also pass any `data.frame` directly — specify column names if they're
non-standard:

```{r eval=FALSE}
manhattan_plot(my_df, chr = "chrom", bp = "position", p = "pval")
```

## Manhattan plot

The most common GWAS visualization:

```{r manhattan-basic}
manhattan_plot(example_gwas)
```

### Highlighting and labeling SNPs

```{r manhattan-labels}
manhattan_plot(
  example_gwas,
  label_top_n = 5,
  genome_wide = 5e-8,
  suggestive = 1e-5
)
```

### Custom colors

```{r manhattan-colors}
manhattan_plot(example_gwas, colors = gwas_palette("vibrant"))
```

## QQ plot

Assess p-value distribution with confidence bands and genomic inflation:

```{r qq-basic}
qq_plot(example_gwas, show_lambda = TRUE, ci = 0.95)
```

## Miami plot

Compare two studies side-by-side:

```{r miami}
# Using same data for demo — in practice, use two different GWAS
miami_plot(
  example_gwas,
  example_gwas,
  top_title = "Discovery",
  bottom_title = "Replication"
)
```

## Locus zoom

Zoom into a specific genomic region:

```{r locus}
locus_plot(
  example_gwas,
  region_chr = 1,
  region_start = 1e6,
  region_end = 50e6
)
```

## Novel visualizations

### Genome-wide p-value heatmap

A compact overview of association signals across the genome:

```{r heatmap}
pvalue_heatmap(example_gwas, bin_size = 10e6, palette = "magma")
```

### Effect-size volcano

See effect sizes and significance simultaneously:

```{r volcano}
volcano_plot(example_gwas, label_top_n = 5)
```

### Circular Manhattan

Circos-style display:

```{r circular}
circular_manhattan(example_gwas, colors = gwas_palette("nature"))
```

## Summary dashboard

A publication-ready multi-panel figure:

```{r summary, fig.height=8}
gwas_summary(example_gwas)
```

## Next steps

- See `vignette("customization")` for themes, palettes, and publication presets
- See `vignette("advanced")` for gene labels on peaks, region highlights,
  top hits tables, multi-trait overlays, enrichment plots, and performance tips
