---
title: "gwasplot: Modern GWAS Visualizations"
author:
  - name: Bartosz Czech
    affiliation: GeneStat
    email: bartosz.w.czech@gmail.com
date: "`r Sys.Date()`"
package: gwasplot
abstract: >
  The gwasplot package provides a comprehensive, ggplot2-based toolkit for
  visualizing genome-wide association study results. It covers the full
  spectrum of GWAS plots — from standard Manhattan and QQ plots to novel
  visualizations like genome-wide heatmaps, enrichment overlays, and
  multi-trait comparisons. All functions return ggplot objects, making them
  fully composable with the tidyverse graphics ecosystem. This vignette
  demonstrates the main functionality with worked examples.
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{gwasplot: Modern GWAS Visualizations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  out.width = "100%",
  dpi = 120
)
```

# Introduction

Genome-wide association studies (GWAS) generate large amounts of summary
statistics that require effective visualization for interpretation and
publication. The standard tools in R (`qqman`, `CMplot`) rely on base R
graphics and offer limited customization. Modern genomics workflows benefit
from a plotting system that integrates with ggplot2 — enabling layered
customization, journal-specific theming, and seamless composition of
multi-panel figures.

`gwasplot` was designed with three goals:

1. **Comprehensive coverage** — all common GWAS plot types in one package,
   plus novel visualizations not available elsewhere.
2. **Publication-ready defaults** — sensible colors, proper axis formatting,
   and built-in journal themes so plots look good without tweaking.
3. **Performance at scale** — smart downsampling handles 10M+ variant datasets
   without manual intervention.

The package reads output from PLINK, REGENIE, GCTA, GEMMA, and any generic
tab-separated file. All plot functions return standard ggplot objects, so
you can add layers, override themes, or combine panels with `patchwork`
exactly as you would with any other ggplot.

# Getting started

## Installation

```{r eval=FALSE}
# From GitHub
pak::pak("bczech/gwasplot")

# Once available on Bioconductor:
# BiocManager::install("gwasplot")
```

## Reading data

The fastest way to load GWAS results is `read_gwas_table()`, which
auto-detects common column naming conventions (CHR/CHROM/#CHROM,
BP/POS/GENPOS, P/PVALUE/LOG10P, etc.):

```{r}
library(gwasplot)
library(ggplot2)

data(example_gwas)
example_gwas
```

The result is a `gwas_data` object — a data.frame with validated columns
and a custom print method showing key QC metrics at a glance.

For specific tools, dedicated readers handle format quirks automatically:

```{r eval=FALSE}
gwas <- read_gwas_table("results.txt")       # auto-detect
gwas <- read_plink_assoc("results.assoc")    # PLINK
gwas <- read_plink_linear("results.assoc.linear")
gwas <- read_regenie("results.regenie")      # REGENIE (handles LOG10P)
gwas <- read_gcta_mlma("results.mlma")       # GCTA
gwas <- read_gemma("results.assoc.txt")      # GEMMA
```

You can also pass any `data.frame` directly — just specify non-standard
column names:

```{r eval=FALSE}
manhattan_plot(my_df, chr = "chrom", bp = "position", p = "pval")
```

# Core plots

## Manhattan plot

The Manhattan plot is the workhorse of GWAS visualization — each point is a
tested variant, plotted by genomic position (x) against its association
strength as -log10(p-value) (y). Chromosomes alternate in color for visual
separation. Dashed lines mark genome-wide significance (5×10⁻⁸) and the
suggestive threshold (10⁻⁵).

```{r manhattan-basic}
manhattan_plot(example_gwas)
```

Label the strongest signals and switch to a different palette:

```{r manhattan-custom}
manhattan_plot(
  example_gwas,
  label_top_n = 5,
  colors = gwas_palette("vibrant")
)
```

## QQ plot

The quantile-quantile plot compares observed p-value distribution against the
expectation under the null hypothesis. Deviation from the diagonal (especially
in the tail) indicates true signal, while early departure suggests population
stratification or technical artifacts. The grey band shows the 95% confidence
interval, and λ_GC quantifies genomic inflation.

```{r qq}
qq_plot(example_gwas, show_lambda = TRUE, ci = 0.95)
```

Stratifying by minor allele frequency helps diagnose whether inflation is
driven by rare variants (common in imputed data) or is uniform:

```{r qq-stratified}
example_gwas$MAF_bin <- cut(
  example_gwas$AF,
  breaks = c(0, 0.05, 0.2, 0.5),
  labels = c("Rare (<5%)", "Low (5-20%)", "Common (>20%)")
)
qq_plot(example_gwas, group = "MAF_bin")
```

## Miami plot

The Miami plot mirrors two Manhattan plots vertically — one pointing up, one
pointing down. This is the standard way to compare discovery and replication
cohorts, or two related traits, on a shared genomic coordinate axis:

```{r miami}
miami_plot(
  example_gwas, example_gwas,
  top_title = "Discovery",
  bottom_title = "Replication"
)
```

## Locus zoom

Once you identify a significant region, a locus plot shows the fine structure
of association within that locus. Optional LD coloring and gene tracks add
biological context:

```{r locus}
locus_plot(example_gwas, region_chr = 1, region_start = 1e6, region_end = 50e6)
```

# Novel visualizations

## Genome-wide p-value heatmap

A compact representation of association signals across the entire genome.
The x-axis is binned genomic position, the y-axis is chromosome, and color
intensity reflects the strongest signal in each bin. This works well as an
overview figure when you want to show genome-wide patterns without
overplotting thousands of points. It also handles very large datasets
efficiently because it pre-aggregates before plotting.

```{r heatmap}
pvalue_heatmap(example_gwas, bin_size = 10e6, palette = "magma")
```

## Effect-size volcano

Unlike the RNA-seq volcano plot (which uses fold change), this plots GWAS
effect size (BETA) against significance. You see at once which signals are
protective vs risk-increasing, and how large the effects are relative to
their statistical evidence. Points can be colored by significance status
or by chromosome:

```{r volcano}
volcano_plot(example_gwas, label_top_n = 5)
```

## Circular Manhattan

The circular (circos-style) Manhattan arranges chromosomes around a ring.
This layout is compact and works particularly well for multi-trait
comparisons where each trait occupies its own ring:

```{r circular, fig.width=8, fig.height=8}
circular_manhattan(example_gwas, colors = gwas_palette("nature"))
```

## Summary dashboard

When you need a single figure that shows the full picture — Manhattan, QQ,
top hits table, and p-value distribution — `gwas_summary()` assembles them
into a multi-panel layout with automatic tags (A, B, C, D) suitable for
supplementary figures:

```{r summary, fig.height=9}
gwas_summary(example_gwas)
```

# Gene annotation and top hits

## Gene labels on peaks

In manuscripts, rs IDs are less informative than gene names.
`manhattan_genes()` takes a gene annotation table and labels each lead SNP
with its nearest gene. In practice, you would extract this table from a GTF
file or via `biomaRt` — here we define a small set manually:

```{r gene-labels}
genes <- data.frame(
  chr = c(1, 2, 5, 7, 11, 15, 19, 22),
  start = c(2e6, 50e6, 80e6, 30e6, 60e6, 40e6, 10e6, 20e6),
  end = c(3e6, 52e6, 85e6, 35e6, 65e6, 45e6, 15e6, 25e6),
  gene = c("MTOR", "NRXN1", "TCF7L2", "CDKAL1", "KCNQ1", "FTO", "LDLR", "COMT")
)

manhattan_genes(example_gwas, genes = genes, gene_p_threshold = 0.001)
```

When peaks are close together, arrow-style annotations avoid label overlap:

```{r gene-arrows}
manhattan_genes(
  example_gwas, genes = genes,
  gene_p_threshold = 0.001,
  arrow = TRUE, label_face = "italic"
)
```

## Top hits table

`top_hits()` extracts independent lead SNPs using window-based clumping
(1 Mb by default), annotates each with an estimated cytoband, and optionally
maps to the nearest gene. The result is a plain data.frame ready for export
to a manuscript table:

```{r top-hits}
top_hits(example_gwas, p_threshold = 0.001, genes = genes, n = 10)
```

## Region highlights

Mark specific genomic regions with a colored band — for example the MHC
(chr6:25-34 Mb), known risk loci from prior GWAS, or candidate regions
from linkage studies:

```{r regions}
plt <- manhattan_plot(example_gwas)

regions <- data.frame(
  chr = c(6, 1),
  start = c(25e6, 5e6),
  end = c(35e6, 30e6),
  label = c("MHC", "1p36")
)

highlight_regions(plt, regions, color = "#FFD700", alpha = 0.2)
```

# Multi-study comparisons

## Multi-trait Manhattan

Comparing GWAS results across traits on a single plot reveals shared
genetic architecture (pleiotropy). `multitrait_manhattan()` overlays
multiple datasets with distinct colors and automatically highlights variants
that are genome-wide significant in more than one trait:

```{r multitrait}
set.seed(123)
trait2 <- example_gwas
trait2$P <- runif(nrow(trait2))^3
trait2$P[sample(nrow(trait2), 10)] <- 10^(-runif(10, 5, 10))

multitrait_manhattan(
  BMI = example_gwas,
  Height = trait2,
  colors = "nature",
  highlight_shared = TRUE
)
```

## Enrichment Manhattan

Functional genomic annotations (enhancers, promoters, eQTLs, open chromatin
regions) help interpret whether GWAS signals cluster in biologically
meaningful parts of the genome. `enrichment_manhattan()` colors variants
by their overlap with user-supplied annotation regions:

```{r enrichment}
annotations <- data.frame(
  chr = c(1, 3, 5, 7, 11, 15, 19),
  start = c(5e6, 20e6, 50e6, 30e6, 60e6, 40e6, 10e6),
  end = c(30e6, 50e6, 120e6, 80e6, 100e6, 80e6, 40e6),
  category = c("Enhancer", "Gene", "eQTL", "Gene",
               "Enhancer", "eQTL", "Gene")
)

enrichment_manhattan(example_gwas, annotations = annotations, palette = "nature")
```

## Circular Manhattan (multi-ring)

The multi-ring layout stacks traits concentrically — each ring is an
independent GWAS, sharing the same angular coordinate system:

```{r circular-multi, fig.width=8, fig.height=8}
circular_manhattan(
  list(BMI = example_gwas, Height = trait2),
  colors = "nature", point_size = 0.6
)
```

# Post-GWAS analysis plots

## PheWAS plot

Phenome-wide association studies test one variant against hundreds or
thousands of phenotypes. The PheWAS plot arranges phenotypes by category
along the x-axis, with -log10(p) on the y-axis. Triangle direction
indicates whether the effect is positive or negative — immediately showing
which phenotype domains are enriched for associations:

```{r phewas, fig.height=5}
phewas_data <- data.frame(
  phenotype = paste0("Pheno_", 1:60),
  p = 10^(-runif(60, 0, 10)),
  category = rep(c("Metabolic", "Immune", "Neurological",
                    "Cardiovascular", "Musculoskeletal", "Other"), 10),
  beta = rnorm(60, 0, 0.3)
)

phewas_plot(phewas_data, label_top_n = 5)
```

## Colocalization locus plot

When a GWAS signal and an eQTL signal overlap at the same locus, a
colocalization plot helps judge visually whether they share the same
causal variant. The two panels share a genomic coordinate axis — if the
peaks line up, colocalization is likely:

```{r coloc}
set.seed(42)
eqtl_data <- example_gwas
eqtl_data$P <- runif(nrow(eqtl_data))^2

coloc_plot(
  example_gwas, eqtl_data,
  region_chr = 1, region_start = 1e6, region_end = 30e6,
  top_title = "GWAS", bottom_title = "eQTL"
)
```

## Fine-mapping credible set

After statistical fine-mapping (SuSiE, FINEMAP), each variant has a
posterior inclusion probability (PIP) indicating how likely it is to be
causal. `finemapping_plot()` encodes PIP as point size and credible set
membership as color — the emerging standard for fine-mapping figures:

```{r finemapping}
fm_data <- example_gwas
fm_data$PIP <- runif(nrow(fm_data))^5
fm_data$PIP[which.min(fm_data$P)] <- 0.92
fm_data$PIP[order(fm_data$P)[2:3]] <- c(0.65, 0.4)
fm_data$credible_set <- NA_integer_
fm_data$credible_set[fm_data$PIP > 0.05] <- 1L

finemapping_plot(
  fm_data,
  region_chr = 1, region_start = 1e6, region_end = 50e6,
  label_pip_above = 0.3
)
```

## Genetic correlation matrix

LD Score Regression (LDSC) estimates genetic correlations between traits.
`genetic_correlation()` displays them as a clustered heatmap with
significance markers — one of the most common figures in cross-trait
genomics papers:

```{r rg-matrix, fig.height=6}
traits <- c("BMI", "Height", "WHR", "T2D", "CAD", "SBP", "HDL")
set.seed(123)
rg <- matrix(0, 7, 7, dimnames = list(traits, traits))
rg[lower.tri(rg)] <- runif(21, -0.6, 0.8)
rg <- rg + t(rg)
diag(rg) <- 1

genetic_correlation(rg, show_values = TRUE, palette = "RdBu")
```

## Genetic architecture

The relationship between minor allele frequency and effect size reveals
a trait's genetic architecture. Polygenic traits (height, BMI) show many
small effects spread across the MAF spectrum. Oligogenic traits have a
few large-effect variants, often at low frequency:

```{r architecture}
architecture_plot(example_gwas, p_threshold = 0.001, label_top_n = 5)
```

# Themes, palettes, and presets

## Color palettes

gwasplot ships with 14 palettes covering different use cases — colorblind-safe
defaults, journal-inspired schemes, and RColorBrewer integrations:

```{r}
gwas_palettes()
gwas_palette("nature", n = 5)
```

Any palette can be passed directly to plot functions via the `colors` argument.
Use `gwas_palette(name, n = k)` to interpolate when you need a specific number
of colors (e.g., for multi-trait plots).

## Journal themes

Each journal has its own style requirements for figure fonts, sizes, and axis
formatting. Built-in themes handle this so you don't need to manually adjust
parameters for each submission:

```{r theme-demo}
manhattan_plot(example_gwas, label_top_n = 3) + theme_nature()
```

Available themes: `theme_nature()`, `theme_science()`, `theme_cell()`,
`theme_plos()`, `theme_presentation()`, `theme_poster()`.

## Presets

A preset bundles together a theme, color palette, and appropriate point sizes
for a given context. This is the fastest way to switch between exploratory
analysis and publication-ready output:

```{r preset}
p <- gwas_preset("publication")
manhattan_plot(example_gwas, colors = p$colors, point_size = p$point_size) + p$theme
```

Available presets: `"publication"`, `"presentation"`, `"poster"`,
`"exploratory"`.

# Performance

For large datasets (common in imputed GWAS with 10-40M variants),
gwasplot applies smart downsampling by default. The algorithm preserves all
significant and near-significant variants while spatially binning the dense
cloud of non-significant points. The resulting plot is visually
indistinguishable from plotting all points, but renders in seconds rather
than minutes:

```{r eval=FALSE}
manhattan_plot(large_gwas)                        # auto (target: 200k points)
manhattan_plot(large_gwas, downsample = FALSE)    # all points (slow)
manhattan_plot(large_gwas, downsample_n = 500000) # custom target
```

# Multi-panel figures

Since every gwasplot function returns a standard ggplot object, you can
compose multi-panel figures using `patchwork` the same way you would with
any other ggplot:

```{r patchwork, fig.height=8}
library(patchwork)

p1 <- manhattan_plot(example_gwas, label_top_n = 3) +
  labs(tag = "A") + theme_nature()
p2 <- qq_plot(example_gwas) +
  labs(tag = "B") + theme_nature()
p3 <- volcano_plot(example_gwas, label_top_n = 5) +
  labs(tag = "C") + theme_nature()

p1 / (p2 | p3) + plot_layout(heights = c(2, 1))
```

# Saving figures

Standard ggplot2 export with dimensions matching journal requirements:

```{r eval=FALSE}
fig <- manhattan_plot(example_gwas, label_top_n = 5) + theme_nature()

# Nature — single column (89 mm width)
ggsave("figure1.pdf", fig, width = 89, height = 55, units = "mm")

# Full-width figure (183 mm)
ggsave("figure1_wide.pdf", fig, width = 183, height = 80, units = "mm")

# High-res PNG for presentations
ggsave("figure1.png", fig, width = 10, height = 5, dpi = 300)
```

# Session info

```{r}
sessionInfo()
```
