---
title: "gwasplot: Modern GWAS Visualizations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{gwasplot: Modern GWAS Visualizations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dpi = 96
)
```

# Getting started

## Installation

```{r eval=FALSE}
pak::pak("bczech/gwasplot")
```

## Reading data

`read_gwas_table()` auto-detects column names from any delimited file:

```{r}
library(gwasplot)
library(ggplot2)

data(example_gwas)
example_gwas
```

Format-specific readers are also available:

```{r eval=FALSE}
gwas <- read_gwas_table("results.txt")
gwas <- read_plink_assoc("results.assoc")
gwas <- read_plink_linear("results.assoc.linear")
gwas <- read_regenie("results.regenie")
gwas <- read_gcta_mlma("results.mlma")
gwas <- read_gemma("results.assoc.txt")
```

Any `data.frame` works directly — just map the columns if needed:

```{r eval=FALSE}
manhattan_plot(my_df, chr = "chrom", bp = "position", p = "pval")
```

# Core plots

## Manhattan plot

```{r manhattan-basic}
manhattan_plot(example_gwas)
```

Label the top hits and adjust colors:

```{r manhattan-custom}
manhattan_plot(
  example_gwas,
  label_top_n = 5,
  colors = gwas_palette("vibrant")
)
```

## QQ plot

P-value distribution with confidence bands and genomic inflation factor:

```{r qq}
qq_plot(example_gwas, show_lambda = TRUE, ci = 0.95)
```

Stratify by MAF to check whether inflation is driven by rare variants:

```{r qq-stratified}
example_gwas$MAF_bin <- cut(
  example_gwas$AF,
  breaks = c(0, 0.05, 0.2, 0.5),
  labels = c("Rare (<5%)", "Low (5-20%)", "Common (>20%)")
)
qq_plot(example_gwas, group = "MAF_bin")
```

## Miami plot

Two GWAS results mirrored vertically — handy for discovery vs replication:

```{r miami}
miami_plot(
  example_gwas, example_gwas,
  top_title = "Discovery",
  bottom_title = "Replication"
)
```

## Locus zoom

Zoom into a region of interest:

```{r locus}
locus_plot(example_gwas, region_chr = 1, region_start = 1e6, region_end = 50e6)
```

# Novel visualizations

## Genome-wide p-value heatmap

Each cell is a genomic bin colored by the minimum p-value it contains.
Handles millions of variants without downsampling because it pre-aggregates:

```{r heatmap}
pvalue_heatmap(example_gwas, bin_size = 10e6, palette = "magma")
```

## Effect-size volcano

BETA vs -log10(p) — see direction and magnitude of effect at a glance:

```{r volcano}
volcano_plot(example_gwas, label_top_n = 5)
```

## Circular Manhattan

Circos-style layout, also supports multiple rings for multi-trait comparison:

```{r circular}
circular_manhattan(example_gwas, colors = gwas_palette("nature"))
```

## Summary dashboard

Multi-panel figure with automatic panel tags (A, B, C, D):

```{r summary, fig.height=8}
gwas_summary(example_gwas)
```

# Gene annotation and top hits

## Gene labels on peaks

`manhattan_genes()` replaces rs IDs with the nearest gene name from a
user-supplied table. In practice you would load this from a GTF file
or via biomaRt — here we define a small set by hand:

```{r gene-labels}
genes <- data.frame(
  chr = c(1, 2, 5, 7, 11, 15, 19, 22),
  start = c(2e6, 50e6, 80e6, 30e6, 60e6, 40e6, 10e6, 20e6),
  end = c(3e6, 52e6, 85e6, 35e6, 65e6, 45e6, 15e6, 25e6),
  gene = c("MTOR", "NRXN1", "TCF7L2", "CDKAL1", "KCNQ1", "FTO", "LDLR", "COMT")
)

manhattan_genes(example_gwas, genes = genes, gene_p_threshold = 0.001)
```

Arrow-style labels work better when peaks are close together:

```{r gene-arrows}
manhattan_genes(
  example_gwas, genes = genes,
  gene_p_threshold = 0.001,
  arrow = TRUE, label_face = "italic"
)
```

## Top hits table

`top_hits()` clumps significant SNPs (keeping one lead per 1 Mb window),
estimates the cytoband, and optionally maps each to the nearest gene:

```{r top-hits}
top_hits(example_gwas, p_threshold = 0.001, genes = genes, n = 10)
```

## Region highlights

Draw a colored band behind a genomic region — useful for marking the MHC,
known GWAS loci, or any region of interest:

```{r regions}
plt <- manhattan_plot(example_gwas)

regions <- data.frame(
  chr = c(6, 1),
  start = c(25e6, 5e6),
  end = c(35e6, 30e6),
  label = c("MHC", "1p36")
)

highlight_regions(plt, regions, color = "#FFD700", alpha = 0.2)
```

# Multi-study comparisons

## Multi-trait Manhattan

Overlay multiple GWAS on one plot. Variants reaching significance in
more than one trait are flagged automatically:

```{r multitrait}
set.seed(123)
trait2 <- example_gwas
trait2$P <- runif(nrow(trait2))^3
trait2$P[sample(nrow(trait2), 10)] <- 10^(-runif(10, 5, 10))

multitrait_manhattan(
  BMI = example_gwas,
  Height = trait2,
  colors = "nature",
  highlight_shared = TRUE
)
```

## Enrichment Manhattan

Color SNPs by whether they fall in annotated functional regions
(enhancers, promoters, eQTLs). Pass a data.frame or a named list:

```{r enrichment}
annotations <- data.frame(
  chr = c(1, 3, 5, 7, 11, 15, 19),
  start = c(5e6, 20e6, 50e6, 30e6, 60e6, 40e6, 10e6),
  end = c(30e6, 50e6, 120e6, 80e6, 100e6, 80e6, 40e6),
  category = c("Enhancer", "Gene", "eQTL", "Gene",
               "Enhancer", "eQTL", "Gene")
)

enrichment_manhattan(example_gwas, annotations = annotations, palette = "nature")
```

## Circular Manhattan (multi-ring)

```{r circular-multi, fig.height=7}
circular_manhattan(
  list(BMI = example_gwas, Height = trait2),
  colors = "nature", point_size = 0.6
)
```

# Themes, palettes, and presets

## Color palettes

14 built-in palettes, all colorblind-safe:

```{r}
gwas_palettes()
gwas_palette("nature", n = 5)
```

## Journal themes

```{r theme-demo}
manhattan_plot(example_gwas, label_top_n = 3) + theme_nature()
```

Available: `theme_nature()`, `theme_science()`, `theme_cell()`,
`theme_plos()`, `theme_presentation()`, `theme_poster()`.

## Presets

A preset bundles a theme, palette, and point sizes:

```{r preset}
p <- gwas_preset("publication")
manhattan_plot(example_gwas, colors = p$colors, point_size = p$point_size) + p$theme
```

Options: `"publication"`, `"presentation"`, `"poster"`, `"exploratory"`.

# Performance

Smart downsampling kicks in automatically above 200k variants. It keeps
every significant SNP and bins the rest — the plot looks identical but
renders much faster:

```{r eval=FALSE}
manhattan_plot(large_gwas)                       # auto
manhattan_plot(large_gwas, downsample = FALSE)   # all points
manhattan_plot(large_gwas, downsample_n = 500000) # custom target
```

# Multi-panel figures

Every function returns a ggplot, so `patchwork` works out of the box:

```{r patchwork, fig.height=8}
library(patchwork)

p1 <- manhattan_plot(example_gwas, label_top_n = 3) +
  labs(tag = "A") + theme_nature()
p2 <- qq_plot(example_gwas) +
  labs(tag = "B") + theme_nature()
p3 <- volcano_plot(example_gwas, label_top_n = 5) +
  labs(tag = "C") + theme_nature()

p1 / (p2 | p3) + plot_layout(heights = c(2, 1))
```

# Saving

```{r eval=FALSE}
fig <- manhattan_plot(example_gwas, label_top_n = 5) + theme_nature()
ggsave("figure1.pdf", fig, width = 89, height = 55, units = "mm")
```

# Session info

```{r}
sessionInfo()
```
